Leaf mass per area, not total leaf area, drives differences in above-ground biomass distribution among woody plant functional types

### 2016-04-26 manuscript.Rnw
more edits

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion.
\begin{itemize}
\item Here we aim to understand differences in biomass distribution between major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm) in terms of underlying traits, in particular the leaf mass per area, a key leaf trait.
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, leaf mass per area, total leaf area, stem basal area and climate on above-ground biomass distribution. 
\item The ratio of leaf mass to above-ground woody mass ($\mfms$) varied strongly between PFTs. We found that $\mfms$, at a given plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation in biomass distribution at a given plant size.
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

The composition of $\mfms$ and $\afms$ into component variables confirmed our hypotheses, that size-related effects in biomass distribution are heavily concentrated in the $\asms$ term within eq. \ref{eq:mfmspart}-\ref{eq:afmspart}. For $\mfms$, $\afms$ and each of the component variables, we estimated the variance explained by individual height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}). We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model. The first model included only plant height, and this variable alone explained >65\% of the variation in both $\mfms$, $\afms$, and $\asms$, but little variation (4 - 11 \%) in LMA and $\afas$.

We found large differences between the three major PFTs in terms of above-ground biomass distribution, as quantified by the ratio $\mfms$ (Fig. 3a), but not in the ratio $\afms$ (Fig. 3c). Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm. Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m). These three size-standardised values of $\mfms$ were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig. 3d), because the differences in $\mfms$ were exactly compensated by differences in LMA.

These patterns in biomass distribution reflected differences in the distribution of components of eq. \ref{eq:mfmspart}-\ref{eq:afmspart}. In particular, LMA was found to differ substantially among PFTs, whereas $\afas$ did not (Fig. 4), which is consistent with the large differences between PFTs in LMA shown in Fig. 3b.  When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.

To confirm the finding that $\afas$ did not vary appreciably between PFTs, we further examined patterns between $\af$ and $\$as$. When these variables are plotted against one another, we see that similar $\af$ is attained at a given $\as$, across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig. 6). We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate, which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001). Nevertheless, the $R^2$ of the relationship only increased from \Sexpr{round(afas$R2a,3)} to \Sexpr{round(afas$R2b,3)}, indicating that while significant, the PFT had little additional predictive power.

@@

Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mfms$ results from variation in LMA, not $\afas$ or $\asms$; and 2) the majority of height-related variation in $\mfms$ results from variation in $\asms$, not LMA or $\afas$. Nevertheless, it is important to note that there is also very large variation between individuals within PFTs (Fig. 4), and within all species as plant species grow form seedlings to adults.

@@

Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate. Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test  hypotheses that leaf and stem biomass distribution is driven by component traits, LMA and $\afas$. We found that at a common plant height, the ratio of leaf to above-ground woody biomass ($\mfms$) varied due to underling differences in LMA but across three PFTs, but that there were no differences among PFTs in $\afas$. As a result, leaf area per unit above-ground woody biomass did not vary between PFTs. We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.

>>>>

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion.
\begin{itemize}
\item Here we aim to understand differences in biomass distribution between major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm) in terms of underlying traits, in particular the leaf mass per area and leaf area per unit stem basal area.
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, leaf mass per area, total leaf area, stem basal area and climate on above-ground biomass distribution. 
\item The ratio of leaf mass to above-ground woody mass ($\mfms$) varied strongly between PFTs. We found that $\mfms$, at a given plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass, or per unit stem basal area. Climate consistently explained very little additional variation in biomass distribution at a given plant size.
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

The composition of $\mfms$ and $\afms$ into component variables confirmed our second hypothesis, that size-related effects in biomass distribution are heavily concentrated in the $\asms$ term within Eqn \ref{eq:mfmspart}-\ref{eq:afmspart}. For $\mfms$, $\afms$ and each of the component variables, we estimated the variance explained by individual height, PFT, and climate variables (MAT or aridity) (Table \ref{table:varparttable}). We did this by sequentially adding terms to a GAM, and calculating the $R^2$ for each fitted model. The first model included only plant height, and this variable alone explained >65\% of the variation in both $\mfms$, $\afms$, and $\asms$, but little variation (4 - 11 \%) in LMA and $\afas$.

We found large differences between the three major PFTs in terms of above-ground biomass distribution, as quantified by the ratio $\mfms$ (Fig. 3a), but not in the ratio $\afms$ (Fig. 3c). Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm. Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data). These three size-standardised values of $\mfms$ were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig. 3d), because the differences in $\mfms$ were compensated by differences in LMA.

These patterns in biomass distribution reflected differences in the distribution of components of Eqn \ref{eq:mfmspart}-\ref{eq:afmspart}. In particular, LMA differed substantially among PFTs, whereas $\afas$ did not (Fig. 4), which is consistent with the large differences between PFTs in LMA shown in Fig. 3b.  When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.

To confirm the finding that $\afas$ did not vary appreciably between PFTs, we further examined patterns between $\af$ and $\as$. When these variables are plotted against one another, we see that similar $\af$ is attained at a given $\as$, across the entire plant size range (with the exception of extremely large trees with \textgreater{} 1m$^2$ stem basal area, for which comparatively little data was available) (Fig. 6). We fit a GAM to $\af$ as a function of $\as$ with and without PFT as a covariate, which showed PFT significantly improved model fit (likelihood ratio test, p \textless{} 0.001). Nevertheless, the $R^2$ of the relationship only increased from \Sexpr{round(afas$R2a,3)} to \Sexpr{round(afas$R2b,3)}, indicating that while significant, the PFT had little additional predictive power.

@@

Taken together, these results demonstrate that 1) the majority of PFT-related variation in $\mfms$ results from variation in LMA, not $\afas$ or $\asms$; and 2) the majority of height-related variation in $\mfms$ results from variation in $\asms$, not LMA or $\afas$. Nevertheless, it is important to note that there is also very large variation between individuals within PFTs (Fig. 4), and within all species as plant species grow from seedlings to adults.

@@

Using the largest database of individual woody plant allometry to date, we studied patterns in above-ground biomass distribution among major woody plant functional types (PFTs), plant size, and climate. Our compilation is the first of its kind that also includes leaf area estimates on the same plants, allowing a direct test of the hypothesis that leaf and stem biomass distribution is driven by component traits, LMA and $\afas$. We found that at a common plant height, the ratio of leaf to above-ground woody biomass ($\mfms$) varied due to underling differences in LMA but across three PFTs, but that there were no differences among PFTs in $\afas$. As a result, leaf area per unit above-ground woody biomass did not vary between PFTs. We found very weak effects of climate (mean annual temperature and aridity) on biomass distribution, so that the vast majority of the variation in biomass distribution was explained by plant size and PFT.
